---
title: "Data Analysis & Visualisation"
description: |
  This page is a showcase of work completed for the GV300 Advanced Quantitative Political Analysis module.
output: distill::distill_article
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r, include=FALSE}
library(stringr)
library(readr)
library(ggplot2)
library(curl)
library(tidyverse)
kidiq <- read_csv("kidiq-2.csv")
```

# Data Organisation and Manipulation
### Manipulating data to create categories for better analysis
- Creating categories for the ordinal variables of mother's work status, mother's IQ, and mother's level of education. 
- The level of education and work status are combined to create a new "advantage" variable.

```{r, echo=TRUE}

# Create meaningful categories
kidiq <- kidiq %>%
  mutate(
    # Create work status labels
    work_status = case_when(
      mom_work == 1 ~ "Not working",
      mom_work == 2 ~ "Part-time",
      mom_work == 3 ~ "Full-time",
      mom_work == 4 ~ "Full-time+"
    ),
    # Create combined advantage categories
    family_advantage = case_when(
      mom_hs == 0 & mom_work <= 2 ~ "Low advantage",
      mom_hs == 0 & mom_work > 2 ~ "Mixed advantage",
      mom_hs == 1 & mom_work <= 2 ~ "Mixed advantage", 
      mom_hs == 1 & mom_work > 2 ~ "High advantage"
    ),
    # Create IQ terciles for grouping
    iq_group = case_when(
      mom_iq <= quantile(mom_iq, 0.33, na.rm = TRUE) ~ "Lower IQ",
      mom_iq <= quantile(mom_iq, 0.67, na.rm = TRUE) ~ "Middle IQ",
      TRUE ~ "Higher IQ"
    ),
    # Create schooling lables
    mom_school = case_when(
      mom_hs == 0 ~ "Not Completed High School",
      mom_hs == 1 ~ "Completed High School"
    )
  )
# Turning iq_group into a factor so they are ordered correctly when creating the
# visualisation
kidiq$iq_group = factor(kidiq$iq_group, levels = c("Lower IQ", "Middle IQ", "Higher IQ"))

```

### Organising data to create visualisations
- Organising mother's IQ and child test score, which are continuous variables, into broad ordinal categories.
- Combining mother's IQ and child test score to identify "surprising cases", where there is a large disparity between mother's IQ and child attainment in test score.

```{r, echo=TRUE}

# Create categories for the analysis

kidiq <- kidiq %>%
  mutate(
    # Create ability categories
    ability_level = case_when(
      mom_iq <= quantile(mom_iq, 0.25, na.rm = TRUE) ~ "Low ability",
      mom_iq <= quantile(mom_iq, 0.75, na.rm = TRUE) ~ "Medium ability",
      TRUE ~ "High ability"
    ),
    # Create outcome categories  
    outcome_level = case_when(
      kid_score <= quantile(kid_score, 0.25, na.rm = TRUE) ~ "Low outcome",
      kid_score <= quantile(kid_score, 0.75, na.rm = TRUE) ~ "Medium outcome",
      TRUE ~ "High outcome"
    ),
    # Identify "surprising" cases
    surprising_case = case_when(
      ability_level == "High ability" & outcome_level == "Low outcome" ~ "High ability, low outcome",
      ability_level == "Low ability" & outcome_level == "High outcome" ~ "Low ability, high outcome",
      TRUE ~ "Expected pattern"
    ),
    
    # Create groups for age of mother
    age_group = case_when(
      mom_age <= 16 ~ "16 or under",
      mom_age >= 17 & mom_age <= 19 ~ "17-19",
      mom_age >= 20 & mom_age <= 22 ~ "20-22",
      mom_age >= 23 & mom_age <= 25 ~ "23-25",
      mom_age >= 26 & mom_age <= 28 ~ "26-28",
      mom_age >= 29 ~ "29 or over",
    )
  )

```

# Data Visaulisations

### Mean Turnout by Country in the 2024 UK General Election
```{r, include=FALSE}
uk_csv <- "https://electionresults.parliament.uk/general-elections/6/candidacies.csv"
uk_raw <- readr::read_csv(uk_csv, show_col_types = FALSE)
names(uk_raw) <- names(uk_raw) %>% tolower() %>% str_replace_all("\\s+", "_")

uk1 <- uk_raw %>%
  filter(!`election_is_by-election`)

uk2 <- uk1 %>%
  select(
    constituency = constituency_name,
    country      = country_name,
    valid_votes  = candidate_vote_count,
    electorate   = electorate
  )

uk3 <- uk2 %>%
  group_by(country, constituency) %>%
  summarise(
    valid_votes = sum(valid_votes, na.rm = TRUE),
    electorate  = max(electorate, na.rm = TRUE),
    .groups = "drop"
  )

uk <- uk3 %>%
  mutate(turnout_registered = 100 * valid_votes / electorate) %>%
  filter(is.finite(turnout_registered), electorate > 0)

turnout_by_country <- uk %>%
  group_by(country) %>%
  summarise(
    mean_turnout   = mean(turnout_registered, na.rm = TRUE),
    median_turnout = median(turnout_registered, na.rm = TRUE),
    n              = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_turnout))
```
```{r}
ukpalette <- c("Wales" = '#295C17', "Northern Ireland" = '#7F5491',"Scotland" = '#3759aa',"England" = '#af2931')
ggplot(turnout_by_country, aes(x = reorder(country, mean_turnout), y = mean_turnout, fill = country)) +
  geom_col(width = 0.7) +  # bar = mean turnout
  scale_y_continuous(limits = c(0, 100),
                     expand = expansion(mult = c(0, 0.05))) + 
  scale_fill_manual(values = ukpalette) +
  coord_cartesian(ylim = c(50, 60)) +
  labs(
    x = "Country",
    y = "Turnout (% of registered voters)"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

### Distribution of Constituencies by Level of Turnout
Most constituencies fall in the "High" turnout category.

```{r, include=FALSE}
turnout_categories <- uk %>%
  mutate(
    category = case_when(
      turnout_registered < 55 ~ "Low",
      turnout_registered < 65 ~ "Medium",
      turnout_registered < 75 ~ "High",
      TRUE ~ "Very High"
    ),
    # Make it a factor so it plots in order
    category = factor(category, 
                     levels = c("Low", "Medium", "High", "Very High"))
  )

# Count by category
category_counts <- turnout_categories %>%
  group_by(category, country) %>%
  summarise(count = n(), .groups = "drop")
```
```{r}
ggplot(category_counts, aes(x = category, y = count, fill = country)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Turnout Category",
    y = "Number of Constituencies",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom")
```

### Impact of Mothers' Age, Education, and Work Status on Child Test Score
#### Structural Factors Generally Stand
They remain the best predictor of child test score outcomes, though there are some surprising cases.

```{r}
p2palette <- c("Expected pattern" = "#555555","High ability, low outcome" = "#ff4040","Low ability, high outcome" = "#009b77")
your_plot_2 <- kidiq %>%
  # Add your data manipulation here
  ggplot(aes(x = mom_iq, y = kid_score, colour = surprising_case, shape = age_group)) +
  # Add your geoms and aesthetics here
    geom_jitter(alpha = 0.8) +
    labs(
      x = "Mother's IQ (proxy for 'ability')",
      y = "Child's Test Score",
      caption = "Data: Children's test scores and mother's characteristics",
      shape = "Age Group of Mother",
      colour = "Surprising Cases"
    ) +
    scale_colour_manual(values = p2palette) +
    scale_shape_discrete() +
    facet_grid(mom_school ~ factor(work_status, levels = c("Full-time+", "Full-time", "Part-time", "Not working"))) +
    theme_grey() +
    theme(
     axis.text.x = element_text(size = 6),
     legend.text = element_text(size = 5),
     legend.title = element_text(size = 6),
     strip.text.y = element_text(size = 8)
     )

# Uncomment to view your plot
print(your_plot_2)
```

# Data Sources
- National Longitudinal Survey of Youth, Kidiq Dataset
- Parliament Election Results, 2024 UK General Election candidate-level data Available at: https://electionresults.parliament.uk/general-elections/6
